Supplemental Materials for ``Universals of word order reflect optimization of grammars for efficient communication''

### 2019-10-15 writeup/pnas-word-order-si.tex
code

@@

\paragraph{Estimating Parseability}
We use a biaffine attention parser architecture \citep{kiperwasser2016simple,zhang2017dependency,dozat2017stanford}. This architecture is remarkably simple: the words of a sentence are encoded into context-sensitive embeddings using bidirectional LSTMs, then a classifier is trained to predict the head for each work. The classifier works by calculating a score for every pair of word embeddings $(w_i, w_j)$, indicating the likelihood that the $j$th word is the head of the $i$th word. This is a highly generic architecture for recovering graph structures from strings, and is a simplification of graph-based parsers which reduce the parsing problem to a minimal spanning tree problem \citep{mcdonald2005nonprojective}.
The parseability of a sentence $u = u_1\dots u_n$ with syntactic structure $t$ is computed as
\begin{equation}
	\log p_\phi(t|u) = \sum_{i=1}^n \log p_\phi(\text{head}_i, \text{label}_i | u, i)
\end{equation}
where $\text{head}_i \in \{\textsc{root}, 1,\dots,n\}$ is the index of the head of $u_i$ in the syntactic structure, and $\text{label}_i$ is its syntactic relation as formalized in UD; $\phi$ denotes the parameters estimated on the training data (see paragraph `Optimization Details').
The overall parseability is estimated as a Monte Carlo estimate on held-out data:
\begin{equation}\label{eq:rpars}
	R_{Pars} := \operatorname{I}[\utterance,\tree] = \sum_{t,u} p(t,u) \log \frac{p_\phi(t|u)}{p(t)} = \frac{1}{|\text{Heldout Data}|} \sum_{t,u \in \text{Heldout Data}} \log \frac{p_\phi(t|u)}{p(t)}
\end{equation}
The constant $p(t)$ only depends on the language (but not on the word order rules), and can thus be ignored when comparing different grammars applied to the same language, and when optimizing grammars for a given language; we therefore do not attempt to explicitly estimate it.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../results/plane/unlexicalized/pareto-plane-perLanguage-lexicalized-mle.pdf}
	\caption[Predictability and Parseability]{Study 1, replication with lexicalized models: Predictability and parseability of 51 languages, for \emph{lexicalized} models, compare Figure~\ref{fig:pareto-per-lang}.}\label{fig:pareto-plane-noPOS}
\end{figure}

>>>>

\subsection{Parseability and Surprisal Metrics for Observed Orders and Extracted Grammars}

We note that the numbers are below the state of the art in NLP.

\begin{table}
\centering
\small{
\begin{tabular}{l||ll|ll|l||ll|ll|l|lllll}
	Language & \multicolumn{5}{c||}{Observed Orders} & \multicolumn{5}{c|}{Extracted Grammars} \\ 
	& UAS & LAS & U.Pars. & L.Pars. & Surp. & UAS & LAS & U.Pars. & L.Pars. & Surp.\\ \hline \hline
\input{../results/plane/results-observed-and-mle.tex }
\end{tabular}
}
	\caption{Parsing and Surprisal metrics for observed orders (left), and for corpora ordered according to extracted real grammars (right). UAS and LAS refer to \emph{Unlabeled} and \emph{Labeled Attachment Scores}, respectively, indicating what fraction of words is assigned the correct head (UAS) or the correct head and relation label (LAS) when choosing heads and labels assigned the highest probability $p_\phi(\text{head}_i, \text{label}_i | u, i)$ (Equation~\ref{eq:pars-obj}) by the parser.
	U.Pars refers to the average value of $- \log p_\phi(\text{head}_i| u, i)$, which is a measure of the difficulty of recovering the raw tree structure, without relation labels. 
	L.Pars refers to the average value of $- \log p_\phi(\text{head}_i, \text{label}_i| u, i)$, measuring the difficulty of recoovering tree structures including relation labels.
	Note that L.Pars corresponds to $\operatorname{H}[\tree|\utterance]$ normalized by the number of words.
	Finally, Surp. refers to the average word-by-word surprisal, which corresponds to the predictability measure $\operatorname{H}[\utterance]$ normalized by the number of words.
	}\label{tab:observed-and-extracted}
\end{table}

@@

\paragraph{Estimating Parseability}
We use a biaffine attention parser architecture \citep{kiperwasser2016simple,zhang2017dependency,dozat2017stanford}. This architecture is remarkably simple: the words of a sentence are encoded into context-sensitive embeddings using bidirectional LSTMs, then a classifier is trained to predict the head for each work. The classifier works by calculating a score for every pair of word embeddings $(w_i, w_j)$, indicating the likelihood that the $j$th word is the head of the $i$th word. This is a highly generic architecture for recovering graph structures from strings, and is a simplification of graph-based parsers which reduce the parsing problem to a minimal spanning tree problem \citep{mcdonald2005nonprojective}.
The parseability of a sentence $u = u_1\dots u_n$ with syntactic structure $t$ is computed as
\begin{equation}
	\log p_\phi(t|u) = \sum_{i=1}^n \log p_\phi(\text{head}_i, \text{label}_i | u, i)
\end{equation}\label{eq:pars-obj}
where $\text{head}_i \in \{\textsc{root}, 1,\dots,n\}$ is the index of the head of $u_i$ in the syntactic structure, and $\text{label}_i$ is its syntactic relation as formalized in UD; $\phi$ denotes the parameters estimated on the training data (see paragraph `Optimization Details').
The overall parseability is estimated as a Monte Carlo estimate on held-out data:
\begin{equation}\label{eq:rpars}
	R_{Pars} := \operatorname{I}[\utterance,\tree] = \sum_{t,u} p(t,u) \log \frac{p_\phi(t|u)}{p(t)} = \frac{1}{|\text{Heldout Data}|} \sum_{t,u \in \text{Heldout Data}} \log \frac{p_\phi(t|u)}{p(t)}
\end{equation}
The constant $p(t)$ only depends on the language (but not on the word order rules), and can thus be ignored when comparing different grammars applied to the same language, and when optimizing grammars for a given language; we therefore do not attempt to explicitly estimate it.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../results/plane/unlexicalized/pareto-plane-perLanguage-lexicalized.pdf}
	\caption[Predictability and Parseability]{Study 1, replication with lexicalized models: Predictability and parseability of 51 languages, for \emph{lexicalized} models, compare Figure~\ref{fig:pareto-per-lang}.}\label{fig:pareto-plane-noPOS}
\end{figure}
